Deriving Hamiltonian mechanics from physics, without passing through Lagrangians.

### 2014-06-15 2014ClassicalHamiltonianDerivation/classicalHamiltonianDerivation.tex
Settled on a notation for energy and extended hamiltonian

\begin{prop}\label{mdofInvariance}
Let $v$ and $w$ be two vectors defined on the tangent space of the manifold identified by two independent degrees of freedom. Let
\begin{align*}
\omega_{\alpha, \beta} =  \left[
  \begin{array}{cc}
    1 & 0 \\
    0 & 1 \\
  \end{array}
\right] \otimes \left[
  \begin{array}{cc}
    0 & 1 \\
    -1 & 0 \\
  \end{array}
\right] =
\left[
  \begin{array}{cccc}
    0 & 1 & 0 & 0 \\
    -1 & 0 & 0 & 0 \\
    0 & 0 & 0 & 1 \\
    0 & 0 & -1 & 0 \\
  \end{array}
\right] \\
\end{align*}
then $v'^{\alpha} \omega_{\alpha, \beta} w'^{\beta}=v^{\alpha} \omega_{\alpha, \beta} w^{\beta}$ under a bijective map.
\end{prop}

@@

\begin{prop}\label{tdofInvariant}
Let $v$ and $w$ be two vectors defined on the tangent space of the extended phase space for one degree of freedom. Let $\alpha$ and $\beta$ be indexes for $t, E, q, p$. Let
\begin{align*}
\omega_{\alpha, \beta} =  \left[
  \begin{array}{cc}
    -1 & 0 \\
    0 & 1 \\
  \end{array}
\right] \otimes \left[
  \begin{array}{cc}
    0 & 1 \\
    -1 & 0 \\
  \end{array}
\right]
= \left[
  \begin{array}{cccc}
    0 & -1 & 0 & 0 \\
    1 & 0 & 0 & 0 \\
    0 & 0 & 0 & 1 \\
    0 & 0 & -1 & 0 \\
  \end{array}
\right] \\
\end{align*}
then $v'^{\alpha} \omega_{\alpha, \beta} w'^{\beta}=v^{\alpha} \omega_{\alpha, \beta} w^{\beta}$ under detrev transformation.
\end{prop}

$\langle T, E \rangle$ are not independent labels from $\langle Q, P \rangle$ as they do not define new states. So they are not necessarily orthogonal in the extended phase space. Looking back at \ref{disclabelsoverr}, cells need to be defined on the plane where $\langle Q, P \rangle$ (maximally) changes: this is not defined on the plane of constant $\langle E, T \rangle$ (they are not orthogonal) where $dq \wedge dp$ is defined, but on the plane perpendicular to constant $\langle Q, P \rangle$ where $dt \wedge dE$ is defined. On that plane we can properly count states and define our invariant.

As in the case of independent d.o.f., we have three planes of conjugate variables forming a right triangle-like relationship with one side invariant.
\begin{align*}
dq^1 \wedge dp^1 + dq^2 \wedge dp^2 = k \\
dt \wedge dE + k = dq \wedge dp
\end{align*}
In the previous case, the right angle was between the two independent degree of freedom. In this case, the right angle is between the invariant and the plane of constant $\langle Q, P \rangle$ where $dt \wedge dE$ is defined. We rewrite it as $dq \wedge dp - dt \wedge dE = k$. Which corresponds to the Minkowski product across d.o.f. and the vector product within. The metric, with a space-like convention, still gives us the cardinality of labels within a degree of freedom.

\begin{prop}\label{tdofConstrain}
The evolution is constrained by $H_{e}=k$, the generating function defined in \ref{genAntisim} generalized for the extended phase space.
\end{prop}

Given N degrees of freedom, phase space is $\mathbb{R}^{2*N}$. Extended phase space is $\mathbb{R}^{2*N + 2}$. Phase space evolved in time is $\mathbb{R}^{2*N + 1}$. Since $H_{e}$ is constant through the evolution, it can serve both as the generating function and as the evolution constrain. By convention, we set $H_{e}=0$.

\begin{prop}\label{tdofHam}
The evolution for time varying multiple degrees of freedom is given by:
\begin{align*}
d_{s}t &= - \partial_{E} H_{e} \\
d_{s}E &= \partial_{t} H_{e} \\
d_{s}q^i &= \partial_{p^i} H_{e} \\
d_{s}p^i &= - \partial_{q^i} H_{e}
\end{align*}
\end{prop}

Expand \ref{genAntisim} with the metric defined in \ref{tdofInvariant}. We recognise Hamilton equations in the extended phase space.

It should not be a surprise that the equations do not mention $c$. In fact, nothing says that all $q^i$ represent space, and with the same unit, and that the laws of motion are invariant in all inertial frames. The only requirement we have is that the areas of each degrees of freedom represent the same measure for labels.

We wrap up with an example. Let $H_{E}= mc^2 + ((p^i)^2 - E^2/c^2) / 2m$. This represents the rest energy of a free particle, the conjugate of which is proper time. Therefore the parametrization of a particle-state is $s=\tau$ while the parametrization for an anti-particle state is $s=-\tau$. If we apply \ref{tdofHam} we have:
\begin{align*}
d_{s}t &= E / mc^2 \\
d_{s}E &= 0 \\
d_{s}q^i &= p^i / m \\
d_{s}p^i &= 0
\end{align*}
For particles, we have:
\begin{align*}
E / c &= m c d_{\tau}t = m U^0 = P^0 \\
p^i &= m d_{\tau}q^i = m U^i = P^i \\
\end{align*}
so we recognize the four-momentum $P^\alpha = {E/c, p^i}$. For anti-particles:
\begin{align*}
- E / c &= m c d_{\tau}t = m U^0 = P^0 \\
- p^i &= m d_{\tau}q^i = m U^i = P^i \\
\end{align*}
we end up with a minus sign between the four-momentum and the conjugate variables $P^\alpha = {-E/c, -p^i}$.

>>>>

\begin{prop}\label{mdofInvariance}
Let $v$ and $w$ be two vectors defined on the tangent space of the manifold identified by two independent degrees of freedom. Let $\alpha$ and $\beta$ be indexes for the state variables $q^1, p^1, q^2, p^2$. Let
\begin{align*}
\omega_{\alpha, \beta} =  \left[
  \begin{array}{cc}
    1 & 0 \\
    0 & 1 \\
  \end{array}
\right] \otimes \left[
  \begin{array}{cc}
    0 & 1 \\
    -1 & 0 \\
  \end{array}
\right] =
\left[
  \begin{array}{cccc}
    0 & 1 & 0 & 0 \\
    -1 & 0 & 0 & 0 \\
    0 & 0 & 0 & 1 \\
    0 & 0 & -1 & 0 \\
  \end{array}
\right] \\
\end{align*}
then $v'^{\alpha} \omega_{\alpha, \beta} w'^{\beta}=v^{\alpha} \omega_{\alpha, \beta} w^{\beta}$ under a bijective map.
\end{prop}

@@

\begin{prop}\label{tdofInvariant}
Let $v$ and $w$ be two vectors defined on the tangent space of the extended phase space for one degree of freedom. Let $\alpha$ and $\beta$ be indexes for the state variables $t, e, q, p$. Let
\begin{align*}
\omega_{\alpha, \beta} =  \left[
  \begin{array}{cc}
    -1 & 0 \\
    0 & 1 \\
  \end{array}
\right] \otimes \left[
  \begin{array}{cc}
    0 & 1 \\
    -1 & 0 \\
  \end{array}
\right]
= \left[
  \begin{array}{cccc}
    0 & -1 & 0 & 0 \\
    1 & 0 & 0 & 0 \\
    0 & 0 & 0 & 1 \\
    0 & 0 & -1 & 0 \\
  \end{array}
\right] \\
\end{align*}
then $v'^{\alpha} \omega_{\alpha, \beta} w'^{\beta}=v^{\alpha} \omega_{\alpha, \beta} w^{\beta}$ under detrev transformation.
\end{prop}

$\langle T, E \rangle$ are not independent labels from $\langle Q, P \rangle$ as they do not define new states. So they are not necessarily orthogonal in the extended phase space. Looking back at \ref{disclabelsoverr}, cells need to be defined on the plane where $\langle Q, P \rangle$ (maximally) changes: this is not the plane of constant $\langle E, T \rangle$ (they are not orthogonal) where $dq \wedge dp$ is defined, but the plane perpendicular to constant $\langle Q, P \rangle$ where $dt \wedge de$ is defined. On that plane we can properly count states and define our invariant.

As in the case of independent d.o.f., we have three planes of conjugate variables forming a right triangle-like relationship with one side invariant.
\begin{align*}
dq^1 \wedge dp^1 + dq^2 \wedge dp^2 = k \\
dt \wedge de + k = dq \wedge dp
\end{align*}
In the previous case, the right angle was between the two independent degree of freedom. In this case, the right angle is between the invariant and the plane of constant $\langle Q, P \rangle$ where $dt \wedge de$ is defined. We rewrite it as $dq \wedge dp - dt \wedge de = k$. Which corresponds to the Minkowski product across d.o.f. and the vector product within. The metric, with a space-like convention, still gives us the cardinality of labels within a degree of freedom.

\begin{prop}\label{tdofHam}
The evolution for time varying multiple degrees of freedom is given by:
\begin{align*}
d_{s}t &= - \partial_{e} \mathcal{H} \\
d_{s}e &= \partial_{t} \mathcal{H} \\
d_{s}q^i &= \partial_{p^i} \mathcal{H} \\
d_{s}p^i &= - \partial_{q^i} \mathcal{H}
\end{align*}
\end{prop}

Expand \ref{genAntisim} with the metric defined in \ref{tdofInvariant}, calling the parameter $s$ instead of $t$ and generator $\mathcal{H}$ instead of $H$. We recognise Hamilton equations in the extended phase space.

It should not be a surprise that the equations do not mention $c$. In fact, nothing says that all $q^i$ represent space, and with the same unit, and that the laws of motion are invariant in all inertial frames. The only requirement we have is that the areas of each degrees of freedom represent the same cardinality for labels.

\begin{prop}\label{tdofConstrain}
The evolution is constrained by $\mathcal{H}=k$, the generating function defined in \ref{genAntisim} generalized for the extended phase space.
\end{prop}

Given N degrees of freedom, phase space is $\mathbb{R}^{2*N}$. Extended phase space is $\mathbb{R}^{2*N + 2}$. Phase space evolved in time is $\mathbb{R}^{2*N + 1}$. Since $\mathcal{H}$ is constant through the evolution, it can serve both as the generating function and as the evolution constrain. By convention, we set $\mathcal{H}=0$.

\textbf{Example}. We wrap up with an example. Let $\mathcal{H}= mc^2 + ((p^i)^2 - E^2/c^2) / 2m$. The non-constant term represents the rest energy of a free particle, the conjugate of which is proper time. Therefore the parametrization of a particle state is $s=\tau$ while the parametrization for an anti-particle state is $s=-\tau$. If we apply \ref{tdofHam} we have:
\begin{align*}
d_{s}t &= E / mc^2 \\
d_{s}E &= 0 \\
d_{s}q^i &= p^i / m \\
d_{s}p^i &= 0
\end{align*}
For particles, we have:
\begin{align*}
E / c &= m c d_{\tau}t = m U^0 = P^0 \\
p^i &= m d_{\tau}q^i = m U^i = P^i \\
\end{align*}
so we recognize the four-momentum $P^\alpha = {E/c, p^i}$. For anti-particles:
\begin{align*}
- E / c &= m c d_{\tau}t = m U^0 = P^0 \\
- p^i &= m d_{\tau}q^i = m U^i = P^i \\
\end{align*}
we end up with a minus sign between the four-momentum and the conjugate variables $P^\alpha = {-E/c, -p^i}$.
